### 2023-06-18 manuscript/chapters/4_locality.qmd
dkfjwen fhjk !!!!!
valentin proofread and comments

The previous chapter gave a detailed description of the mappings $S : [0,1]^2 \mapsto [0,1]$, as well as a few numerical considerations. We found that mappings that yield equivalent dynamics crucially seem to have in common the property of *locality*, that is that they conserve the regularity between the domain and the image, even if the image seems to have a fractal structure. Intuitively, we saw that the locality seems to emerge from the mapping having overwhelmingly small variations.

In this chapter, we will further explore these results, and attempt to formulate a measure of this "locality". We start by showing that bijectivity and measurability are sufficient conditions for equivalent dynamics. Then, we apply the locality measure to the special cases of the Column and Z-mappings, and show how this notion can discriminate between the two. Finally, we show that the formulated measure is sufficient to ensure the numerical estimation of the integral on the fractal structure converges to the analytical integral; therefore, such embeddings are feasible to find and simulate numerically.

@@

In the previous chapter, we introduced Cantor and Peano functions, and argued that if have to choose between the two, then the non-injective Peano functions are preferable, because, by Netto's theorem, they are the only ones that can be space-filling and reach all the square populations. Despite this, let us make the simplifying assumption that $S$ is a bijective function. However, to not break Netto's theorem, we do not make any assumptions regarding its continuity.

We furthermore assume that the mapping $S$ is measurable. This is not a strong assumption, since the limit of pointwise convergent measurable functions is also measurable. In our case, we have written the finite-$n$ expansions $S^n$, which by being sums of piecewise constant functions, are also measurable. Measurability of $S$ allows us to write $\lambda \circ S^{-1}$, since by definition the preimage of a measurable function is measurable.

@@

In the last line, we have used the bijectivity of $S$. The Lebesgue measure $\lambda : [0,1]^2 \mapsto \R^+$ simply plays the role of a uniform distribution on $[0,1]^2$, and we simply write

@@

[TODO : uniqueness of the solution ?]

@@

### The total variation attempt

@@

[^tv]:
  This definition is slightly different from other definitions of total variation, which involve a supremum over all partitions $P = \{0 \leq \alpha_1 < \cdots < \alpha_{N_p} \leq 1\}$.
  <!-- The total variation is then defined as
  $$
  \mathrm{TV}(S^{-1}) = \sup_{P} \sum_{i=1}^{N_P} \norm{S^{-1}(\alpha_i) - S^{-1}(\alpha_{i-1})}_1.
  $$ -->
  We don't use the more complex definition, because our version of total variation is just used for illustration purposes. Additionally, we make use of the $\ell^1$ norm because it lends itself to the nice geometric intuition of "projecting the jumps" on the axes. This choice of the norm is irrelevant, however, because analytically all norms are equivalent up to a constant.

@@

<!-- In some sense, having the constraint that the mapping traces a path in the bounded space $[0,1]^2$ that can go through each point of the grid only once leaves little flexibility. For fixed $n$, there are only finitely many "large jumps" the mapping can make, and the large number of "small jumps" still add up, so that the total variation still grows at approximately the same rate. -->

@@

When we do the downsampling of the mappings, we expressed that the locality helps by bounding the variation inside of each averaged bin. Therefore, what we would like to express, is that on average, the variation inside of each bin before downsampling is small. @fig-sketch-binned-variation gives an intuition of this. Even though the $[0,1]$ embedding might be discontinuous in places, on average nearby neurons should exhibit the same activity levels.

@@

@@

TODO : I have removed the definition involving the sup over all permutations, because we don't end up using it. Still mention for completeness ?

@@

$$
\begin{aligned}
Z^{-1}_1(\alpha) &= 0.\underbrace{i_{n-1} i_{n-3}}_\text{$n/2$ bits} \cdots i_1 b_n b_{n+2} \cdots b_{2m-1} \\
Z^{-1}_2(\alpha) &= 0.\underbrace{i_{n-2} i_{n-4}}_\text{$n/2$ bits} \cdots i_0 b_{n+1} b_{n+3} \cdots b_{2m}
\end{aligned}
$$

@@

$$
\begin{aligned}
Z^{-1}_1(\alpha) &= 0.\underbrace{i_{n-1} i_{n-3}}_\text{$(n+1)/2$ bits} \cdots i_0 b_n b_{n+2} \cdots b_{2m-1} \\
Z^{-1}_2(\alpha) &= 0.\underbrace{i_{n-2} i_{n-4}}_\text{$(n-1)/2$ bits} \cdots i_1 b_{n+1} b_{n+3} \cdots b_{2m}
\end{aligned}
$$

@@

In @sec-coarse-graining we discussed how we approximate the fractal mappings numerically, and introduced the downsampling procedure, in which we average the populations in segments of length $2^{-n}$, and end up with $2^n$ "binned" segment populations, compared to the $4^n$ we started with.

By doing this, we were able to numerically estimate the integral on the one-dimensional fractal space, and showed that local mappings exhibited the property that they conserved the dynamics between two spaces of different dimensions. @sec-Vn introduced the analytical metric $V_n$ of this locality. Assuming that $V_n \xrightarrow{n\to\infty} 0$, is this enough to prove that the numerical estimation of the integral goes to its analytical value?

The intuition behind the proof of convergence is that the variation, coming from the coarse-graining, inside each numerical bin vanishes as we increase the number of iterations $n$, even though the structure becomes fractal.

@@

Let us first recall the numerical and analytical integral. We leave a $\tilde w(\cdot, \beta)$ as an implicit function of the first argument $\alpha$. 
$$
\begin{aligned}
\mathrm{NI}_n &= \frac{1}{2^n} \sum_{i=1}^{2^n} \tilde w(\cdot, \beta_i) \phi(\tilde h_i(t)) \\
\mathrm{AI} &= \int_{[0,1]} \tilde w(\cdot, \beta) \phi(\tilde h(\beta, t)) \left[\lambda \circ S^{-1}\right](\d \beta)
\end{aligned}
$$

We wish to prove that $|\mathrm{NI}_n - \mathrm{AI}| \xrightarrow{n \to \infty} 0$, and implicitly, that this is valid for all $\alpha \in [0,1]$.

@@

$$
\begin{aligned}
|\mathrm{NI}_n - \mathrm{AI}| &= \left| \frac{1}{2^n} \sum_{i=1}^{2^n} \tilde w(\cdot, \beta_i) \phi(\tilde h_i(t)) - \int_{0}^{1} \tilde w(\cdot, \beta) \phi(\tilde h(\beta, t)) \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
&= \left| \frac{1}{2^n} \sum_{i=1}^{2^n} \tilde w(\cdot, \beta_i) \phi(\tilde h_i(t)) - \sum_{i=1}^{2^n} \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \tilde w(\cdot, \beta) \phi(\tilde h(\beta, t)) \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
&= \left| \sum_{i=1}^{2^n} \frac{1}{2^n} \tilde w(\cdot, \beta_i) \phi(\tilde h_i(t)) - \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \tilde w(\cdot, \beta) \phi(\tilde h(\beta, t)) \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
&\leq \sum_{i=1}^{2^n} \left| \frac{1}{2^n} \tilde w(\cdot, \beta_i) \phi(\tilde h_i(t)) - \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \tilde w(\cdot, \beta) \phi(\tilde h(\beta, t)) \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
\end{aligned}
$$

We now want to prove that evaluating the integrand at $\beta_i = \tfrac{i-1}{2^n}$ gives a good approximation of the integral. Because $S$ is measurable and bijective, we have $\int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \left[\lambda \circ S^{-1}\right](\d \beta) = \tfrac{1}{2^n}$, and we write

$$
\begin{aligned}
|\mathrm{NI}_n - \mathrm{AI}| &\leq \sum_{i=1}^{2^n} \left| \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \Delta(\beta) \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
\text{where}\ \Delta(\beta) &= \tilde w(\cdot, \beta_i) \phi(\tilde h_i(t)) - \tilde w(\cdot, \beta) \phi(\tilde h(\beta, t))\\
&= w_U(\cdot, S^{-1}(\beta_i)) \phi(h_U(S^{-1}(\beta_i), t)) - w_U(\cdot, S^{-1}(\beta)) \phi(h_U(S^{-1}(\beta), t))
\end{aligned}
$$

The main source of variance inside the integrand $\Delta(\beta)$ is the term $S^{-1}$, since $w_U$ and $h_U$ are "nice" functions. We express this regularity by the added assumption that $w_U(\cdot, \vec u) \phi(h_U(t, \vec u))$ is Lipschitz in $\vec u$, with Lipschitz constant $L(\cdot)$, and $\norm{\cdot}$ the corresponding norm on $[0,1]^2$. We argue this is not a strong assumption, because this is just expressing the regularity of the integrand in the integral over $[0,1]^2$.

@@

Applying the Lipschitz property yields the following inequality on the integrand:

$$
\begin{aligned}
\Delta(\beta) &= w_U(\cdot, S^{-1}(\beta_i)) \phi(h_U(S^{-1}(\beta_i), t)) - w_U(\cdot, S^{-1}(\beta)) \phi(h_U(S^{-1}(\beta), t)) \\
&\leq L(\cdot) \norm{S^{-1}(\beta_i) - S^{-1}(\beta)}
\end{aligned}
$$

@@

$$
\begin{aligned}
|\mathrm{NI}_n - \mathrm{AI}| &\leq L(\cdot) \sum_{i=1}^{2^n} \left| \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \norm{S^{-1}(\beta_i) - S^{-1}(\beta)} \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
&\leq L(\cdot) \sum_{i=1}^{2^n} \left| \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \sup_{\beta \in \left[\tfrac{i-1}{2^n}, \tfrac{i}{2^n}\right]} \norm{S^{-1}(\beta_i) - S^{-1}(\beta)} \left[\lambda \circ S^{-1}\right](\d \beta^\prime) \right| \\
&= L(\cdot) \frac{1}{2^n} \sum_{i=1}^{2^n} \sup_{\beta \in \left[\tfrac{i-1}{2^n}, \tfrac{i}{2^n}\right]} \norm{S^{-1}(\beta_i) - S^{-1}(\beta)} \\
&\leq L(\cdot) \frac{1}{2^n} \sum_{i=1}^{2^n} \sup_{\beta, \beta^\prime \in \left[\tfrac{i-1}{2^n}, \tfrac{i}{2^n}\right]^2} \norm{S^{-1}(\beta^\prime) - S^{-1}(\beta)} \\
&= L(\cdot) V_n \\
&\xrightarrow{n \to \infty} 0
\end{aligned}
$$

>>>>

The previous chapter gave a detailed description of the mappings $S : [0,1]^2 \mapsto [0,1]$, as well as a few numerical considerations. We found that mappings that yield equivalent dynamics crucially seem to have in common the property of *locality* (from 1D to 2D), even if the image seems to have a fractal structure. Intuitively, we saw that the locality seems to emerge from the mapping having for the most part small discontinuities.

In this chapter, we will further explore these results, and attempt to formulate a measure of this "locality". As a preamble, we show that bijectivity and measurability are sufficient conditions for  the 2D and 1D neural field dynamics to be equivalent. Then, we apply our measure of "locality" to the special cases of the Column and Z-mappings, and show how this notion can discriminate between the two mappings. Finally, we show that our measure of locality is sufficient to ensure that the numerical estimation of the integral converges to the analytical integral; therefore, such embeddings can be simulated numerically.

@@

<!-- In the previous chapter, we introduced Cantor and Peano functions, and argued that if have to choose between the two, then the non-injective Peano functions are preferable, because, by Netto's theorem, they are the only ones that can be space-filling and reach all the square populations. -->
<!-- Let us make the assumption that $S$ is a bijective function. However, to not break Netto's theorem, we do not make any assumptions regarding its continuity.

Nicole: let's not mention Netto's theorem here. It applies to S^{-1}, and it becomes difficult to reason about when we talk about S. I don't think the proof works strictly in our case, because the Z-mapping is not a bijection, but let us assume uwuwuwu
-->

Let us assume that the mapping $S$ is bijective and measurable. This is not a strong assumption, since the limit of pointwise convergent measurable functions is also measurable. In our case, we have written the finite-$n$ expansions $S^n$, which, by being sums of piecewise constant functions, are also measurable. The measurability of $S$ allows us to write $\lambda \circ S^{-1}$, the pushforward measure (or image measure) of $\lambda$ under the mapping $S$.

@@

In the last line, we used a basic property of the pushfoward measure (see @Bogachev2007, Theorem 3.6.1, p. 190). The Lebesgue measure $\lambda : [0,1]^2 \mapsto \R^+$ simply plays the role of a uniform distribution on $[0,1]^2$, and we simply write

@@

We do not address the question of well-posedness of the neural field equations in this thesis; for proofs of the well-posedness of neural field equations, we refer the reader to [@faugeras2008absolute;@faugeras2009persistent;@veltz2009localglobal].

<!-- TODO : we may argue that the Z-curve is not bijective. This complicates the proof. Though, the pre-image of (v_1, v_2) is composed of at most 4 points, and the measurability compensates. -->

@@

<!-- ### The total variation attempt

@@

[^tv]:
  This definition is slightly different from other definitions of total variation, which involve a supremum over all partitions $P = \{0 \leq \alpha_1 < \cdots < \alpha_{N_p} \leq 1\}$.
  <!-- The total variation is then defined as
  $$
  \mathrm{TV}(S^{-1}) = \sup_{P} \sum_{i=1}^{N_P} \norm{S^{-1}(\alpha_i) - S^{-1}(\alpha_{i-1})}_1.
  $$
  We don't use the more complex definition, because our version of total variation is just used for illustration purposes. Additionally, we make use of the $\ell^1$ norm because it lends itself to the nice geometric intuition of "projecting the jumps" on the axes. This choice of the norm is irrelevant, however, because analytically all norms are equivalent up to a constant.

@@

In some sense, having the constraint that the mapping traces a path in the bounded space $[0,1]^2$ that can go through each point of the grid only once leaves little flexibility. For fixed $n$, there are only finitely many "large jumps" the mapping can make, and the large number of "small jumps" still add up, so that the total variation still grows at approximately the same rate. -->

@@

With this view, we understand that the coarse-grained performed in @sec-coarse-graining was making use of that fact that locality helps by bounding the variation inside of each averaged bin. Therefore, we would like for mathematically formulate that on average, the variation inside of each bin before coarse-graining is small. @fig-sketch-binned-variation gives an intuition of this. Even though the $[0,1]$ embedding might be discontinuous in places, on average nearby neurons should exhibit the same activity levels.

@@

<!-- Nicole: V_n -> 0 is implied by continuity -->

@@

<!-- TODO : I have removed the definition involving the sup over all permutations, because we don't end up using it. Still mention for completeness ?

Valentin: For the master thesis, it is ok to not mention it.
-->

@@

$$
\begin{aligned}
Z^{-1}_1(\alpha) &= 0.\underbrace{i_{n-1} i_{n-3} \cdots i_1}_\text{$n/2$ bits} b_n b_{n+2} \cdots b_{2m-1} \\
Z^{-1}_2(\alpha) &= 0.\underbrace{i_{n-2} i_{n-4} \cdots i_0}_\text{$n/2$ bits} b_{n+1} b_{n+3} \cdots b_{2m}
\end{aligned}of this locality
$$

@@

$$
\begin{aligned}
Z^{-1}_1(\alpha) &= 0.\underbrace{i_{n-1} i_{n-3} \cdots i_0}_\text{$(n+1)/2$ bits} b_n b_{n+2} \cdots b_{2m-1} \\
Z^{-1}_2(\alpha) &= 0.\underbrace{i_{n-2} i_{n-4} \cdots i_1}_\text{$(n-1)/2$ bits} b_{n+1} b_{n+3} \cdots b_{2m}
\end{aligned}
$$

@@

<!-- In @sec-coarse-graining we discussed how we approximate the fractal mappings numerically, and introduced the coarse-graining procedure, in which we average the populations in segments of length $2^{-n}$, and end up with $2^n$ "binned" segment populations, compared to the $4^n$ we started with.

By doing this, we were able to numerically estimate the integral on the one-dimensional fractal space, and showed that local mappings exhibited the property that they conserved the dynamics between two spaces of different dimensions. @sec-Vn introduced the analytical metric $V_n$ of this locality. Assuming that $V_n \xrightarrow{n\to\infty} 0$, is this enough to prove that the numerical estimation of the integral goes to its analytical value? -->

In @sec-Vn, we have introduced the average variation metric $V_n$. Assuming that $V_n \xrightarrow{n\to\infty} 0$ (which expresses the locality property), does this imply that the numerical estimation of the integral in the neural field equation in $[0,1]$ converges to its true value?

The intuition behind the proof of convergence is that the variation, coming from the coarse-graining, inside each numerical bin vanishes as we increase the number of iterations $n$, even though the the neural field $\tilde{h}$ and the connectivity kernel $\tilde{w}$ become fractal.

@@

Let us first recall the numerical and analytical integral. For any $\alpha \in [0,1]$, let us define:
$$
\begin{aligned}
\mathrm{NI}_n &= \frac{1}{2^n} \sum_{i=1}^{2^n} \tilde w(\alpha, \beta_i) \phi(\tilde h_i(t)) \\
\mathrm{AI} &= \int_{[0,1]} \tilde w(\alpha, \beta) \phi(\tilde h(\beta, t)) \left[\lambda \circ S^{-1}\right](\d \beta).
\end{aligned}
$$

We wish to prove that $|\mathrm{NI}_n - \mathrm{AI}| \xrightarrow{n \to \infty} 0$.

@@

$$
\begin{aligned}
|\mathrm{NI}_n - \mathrm{AI}| &= \left| \frac{1}{2^n} \sum_{i=1}^{2^n} \tilde w(\alpha, \beta_i) \phi(\tilde h_i(t)) - \int_{0}^{1} \tilde w(\alpha, \beta) \phi(\tilde h(\beta, t)) \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
&= \left| \frac{1}{2^n} \sum_{i=1}^{2^n} \tilde w(\alpha, \beta_i) \phi(\tilde h_i(t)) - \sum_{i=1}^{2^n} \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \tilde w(\alpha, \beta) \phi(\tilde h(\beta, t)) \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
&= \left| \sum_{i=1}^{2^n} \frac{1}{2^n} \tilde w(\alpha, \beta_i) \phi(\tilde h_i(t)) - \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \tilde w(\alpha, \beta) \phi(\tilde h(\beta, t)) \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
&\leq \sum_{i=1}^{2^n} \left| \frac{1}{2^n} \tilde w(\alpha, \beta_i) \phi(\tilde h_i(t)) - \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \tilde w(\alpha, \beta) \phi(\tilde h(\beta, t)) \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
\end{aligned}
$$

We now want to prove that evaluating the integrand at $\beta_i = \tfrac{i-1}{2^n}$ gives a good approximation of the integral. Let us assume that $S$ is the Z-mapping. Then, $S$ is measure-preserving (if $\lambda$ is the Lebesgue measure on $[0,1]^2$, then $\lambda \circ S^{-1}$ is the Lebesgue measure on $[0,1]$). This implies that we have $\int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \left[\lambda \circ S^{-1}\right](\d \beta) = \tfrac{1}{2^n}$, and we write

$$
\begin{aligned}
|\mathrm{NI}_n - \mathrm{AI}| &\leq \sum_{i=1}^{2^n} \left| \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \Delta(\beta) \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
\text{where}\ \Delta(\beta) &= \tilde w(\alpha, \beta_i) \phi(\tilde h_i(t)) - \tilde w(\alpha, \beta) \phi(\tilde h(\beta, t))\\
&= w_U(\alpha, S^{-1}(\beta_i)) \phi(h_U(S^{-1}(\beta_i), t)) - w_U(\alpha, S^{-1}(\beta)) \phi(h_U(S^{-1}(\beta), t))
\end{aligned}
$$

The main source of variance inside the integrand $\Delta(\beta)$ is the term $S^{-1}$, since $w_U$ and $h_U$ are "nice" functions. We express this regularity by the added assumption that $w_U(\alpha, \vec u) \phi(h_U(t, \vec u))$ is Lipschitz in $\vec u$, with Lipschitz constant $L(\alpha)$, and $\norm{\alpha}$ the corresponding norm on $[0,1]^2$. We argue this is not a strong assumption, because this is just expressing the regularity of the integrand in the integral over $[0,1]^2$.

@@

Applying the Lipschitz assumption yields the following inequality on the integrand:

$$
\begin{aligned}
\Delta(\beta) &= w_U(\alpha, S^{-1}(\beta_i)) \phi(h_U(S^{-1}(\beta_i), t)) - w_U(\alpha, S^{-1}(\beta)) \phi(h_U(S^{-1}(\beta), t)) \\
&\leq L(\alpha) \norm{S^{-1}(\beta_i) - S^{-1}(\beta)}
\end{aligned}
$$

@@

$$
\begin{aligned}
|\mathrm{NI}_n - \mathrm{AI}| &\leq L(\alpha) \sum_{i=1}^{2^n} \left| \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \norm{S^{-1}(\beta_i) - S^{-1}(\beta)} \left[\lambda \circ S^{-1}\right](\d \beta) \right| \\
&\leq L(\alpha) \sum_{i=1}^{2^n} \left| \int_{\tfrac{i-1}{2^n}}^{\tfrac{i}{2^n}} \sup_{\beta \in \left[\tfrac{i-1}{2^n}, \tfrac{i}{2^n}\right]} \norm{S^{-1}(\beta_i) - S^{-1}(\beta)} \left[\lambda \circ S^{-1}\right](\d \beta^\prime) \right| \\
&= L(\alpha) \frac{1}{2^n} \sum_{i=1}^{2^n} \sup_{\beta \in \left[\tfrac{i-1}{2^n}, \tfrac{i}{2^n}\right]} \norm{S^{-1}(\beta_i) - S^{-1}(\beta)} \\
&\leq L(\alpha) \frac{1}{2^n} \sum_{i=1}^{2^n} \sup_{\beta, \beta^\prime \in \left[\tfrac{i-1}{2^n}, \tfrac{i}{2^n}\right]^2} \norm{S^{-1}(\beta^\prime) - S^{-1}(\beta)} \\
&= L(\alpha) V_n \\
&\xrightarrow{n \to \infty} 0
\end{aligned}
$$
